### 2013-07-17 Docs/Thesis/conclusions-and-further-work.tex
thesis updates

Many classes of reactive application would benefit from a ``dynamic collections'' combinator
similar to {\tt pSwitch} in Yampa. Such a combinator allows a collection of signal functions
to be evaluated as one signal function, with addition or removal of signal functions instead
of the total replacement given by the {\tt switch} combinator. This is useful, for instance,
when simulating objects for games or computer models, as the behavior of each object can be
modeled as a signal function, and these signal functions can be added to and removed from the
collection. An attempt was made to add such a function to TimeFlies. Currently,
it is not clear what the type of such a combinator would be in a system
employing signal vectors, or how the "routing" of inputs to individual signal
functions in the collection would be specified.

>>>>

Many classes of reactive application would benefit from a ``dynamic collections'' combinator
similar to {\tt pSwitch} in Yampa. Such a combinator allows a collection of signal functions
to be evaluated as one signal function, with addition or removal of signal functions instead
of the total replacement given by the {\tt switch} combinator. This is useful, for instance,
when simulating objects for games or computer models, as the behavior of each object can be
modeled as a signal function, and these signal functions can be added to and removed from the
collection. An attempt was made to add such a function to TimeFlies, but it is not clear what 
the type of such a combinator would be in a system employing signal vectors, or how the
"routing" of inputs to individual signal functions in the collection would be specified.

### 2013-07-17 Docs/Thesis/implementation-signalfunctions.tex
thesis updates

\begin{figure}
\begin{code}
data SVSample sv where
  SVSample      ::    a
                   -> SVSample (SVSignal a)
  SVSampleEvent ::    SVSample (SVEvent a)
  SVSampleEmpty ::    SVSample SVEmpty
  SVSampleBoth  ::    SVSample svLeft
                   -> SVSample svRight
                   -> SVSample (SVAppend svLeft svRight)
\end{code}
\hrule
\caption{Datatype for signal samples.}
\label{figure:signal_sample_datatype}
\end{figure}

@@

\begin{figure}
\begin{code}
data SVOccurrence sv where
  SVOccurrence ::    a
                  -> SVOccurrence (SVEvent a)
  SVOccLeft    ::    SVOccurrence svLeft
                  -> SVOccurrence (SVAppend svLeft svRight)
  SVOccRight   ::    SVOccurrence svRight 
                  -> SVOccurrence (SVAppend svLeft svRight)
\end{code}
\hrule
\caption{Datatype for event occurrences.}
\label{figure:event_occurrence_datatype}
\end{figure}

@@

\begin{figure}
\begin{code}
data SVDelta sv where
  SVDeltaSignal  ::    a
                    -> SVDelta (SVSignal a)
  SVDeltaNothing ::    SVDelta sv
  SVDeltaBoth    ::    SVDelta svLeft
                    -> SVDelta svRight
                    -> SVDelta (SVAppend svLeft svRight)
\end{code}
\hrule
\caption{Datatype for signal updates.}
\label{figure:signal_update_datatype}
\end{figure}

@@

\begin{figure}
\begin{code}
data Initialized

data NonInitialized

data SF init svIn svOut where
  SF     ::    (SVSample svIn 
                  -> (SVSample svOut,
                      SF Initialized svIn svOut)) 
            -> SF NonInitialized svIn svOut
  SFInit ::    (Double 
                  -> SVDelta svIn
                  -> (SVDelta svOut,
                      [SVOccurrence svOut],
                      SF Initialized svIn svOut)) 
            -> (SVOccurrence svIn
                  -> ([SVOccurrence svOut],
                      SF Initialized svIn svOut))
            -> SF Initialized svIn svOut
\end{code}
\hrule
\caption{Datatype and empty types for signal functions.}
\label{figure:signal_function_datatype}
\end{figure}

@@

composeInit ::     SF Initialized svIn svBetween
                -> SF Initialized svBetween svOut
                -> SF Initialized svIn svOut
composeInit (SFInit dtCont1 inputCont1) sf2@(SFInit dtCont2 inputCont2) =
  SFInit
    (\dt sigDelta -> 
       let (sf1MemOutput, sf1EvtOutputs, sf1New) = dtCont1 dt sigDelta
           (sf2MemOutput, sf2EvtOutputs, sf2New) = dtCont2 dt sf1MemOutput
           (sf2EvtEvtOutputs, sf2Newest) = applySF sf2New sf1EvtOutputs
       in (sf2MemOutput,
           sf2EvtOutputs ++ sf2EvtEvtOutputs,
           composeInit sf1New sf2Newest)
    )
    (\evtOcc -> 
      let (sf1Outputs, newSf1) = inputCont1 evtOcc
          (sf2FoldOutputs, newSf2) = applySF sf2 sf1Outputs
      in (sf2FoldOutputs, composeInit newSf1 newSf2)   
    )

applySF ::    SF Initialized svIn svOut
           -> [SVOccurrence svIn]
           -> ([SVOccurrence svOut],
               SF Initialized svIn svOut)
applySF sf indices =
  foldr (\evtOcc (changes, SFInit _ changeCont) ->
           let (newChanges, nextSF) = changeCont evtOcc
               in (newChanges ++ changes, nextSF))
        ([], sf)
        indices
\end{code}
\hrule
\caption{Implementation of serial composition.}
\label{figure:serial_composition_implementation}
\end{figure}

The switch combinator is the means of introducing reactivity into a signal
function. This combinator allows a signal function to replace itself by
producing an event occurrence. The combinator wraps a signal function, and 
observes an event on the right side of the output signal vector. At the first
occurrence of the event, the signal function carried by the occurrence replaces
the signal function.

The switch combinator stores the input sample provided during initialization,
and updates it with the input signal updates. When the wrapped signal function
produces an occurrence carrying a new signal function, that signal function is
initialized with the stored input sample. It is then "wrapped" by another
function which closes over its output sample, and outputs the sample as a signal
update as the next time step. After this, it acts as the new signal function.
This wrapping has some performance implications, which are discussed

@@

\begin{figure}
\begin{code}
-- T.swap is imported from Data.Tuple
swap :: (sv1 :^: sv2) :~> (sv2 :^: sv1)
swap =
  SF ((, swapInit) . 
      uncurry combineSamples .
      T.swap . splitSample)

swapInit :: SF Initialized (SVAppend sv1 sv2) (SVAppend sv2 sv1)
swapInit =
  SFInit (flip (const .
                (, [], swapInit) .
                uncurry combineDeltas .
                T.swap . splitDelta))
          (\evtOcc ->
             (case chooseOccurrence evtOcc of
                Left lOcc  -> [occRight lOcc]
                Right rOcc -> [occLeft rOcc], swapInit))

@@

The implementation of the joining combinators is simple. The {\tt union}
combinator simply passes along every event occurrence it receives on either
input, stripping off the left and right combinator. This is acceptable since we
do not insist on a total ordering of events, or an event time resolution greater
than the sampling rate. The {\tt combineSignals} combinator maintains the values
of both signals, and applies the combination function whenever one is updated.
The {\tt capture} combinator maintains the input signal value, and adds it to
each event occurrence.

@@

The implementation strategy leaves room for optimizations. In particular, an
additional constructor for time-independent signal functions would allow
portions of a signal function to forgo evaluation during time steps unless they
had signal updates. Optimizations in the style of Yampa, observed by keeping
an updated AST for the signal function and pattern-matching on it when switching,
might further improve performance. In particular, collapsing nested or
serially-composed versions of the {\tt switchWait} step when switching would
remove at least some of the observed dependence of performance on sampling rate.
Nevertheless, this implementation performs quite well as it currently exists, as
we demonstrate in Chapter~\ref{chapter:Evaluation_and_Comparisons}.

>>>>

\begin{figure}
\begin{code}
data SVSample sv where
  -- Store a value at an SVSignal leaf
  SVSample      ::    a
                   -> SVSample (SVSignal a)
  -- Empty constructor for event leaves
  SVSampleEvent ::    SVSample (SVEvent a)
  -- Empty constructor for empty leaves
  SVSampleEmpty ::    SVSample SVEmpty
  -- Construct a sample from two samples, typed
  --  with the SVAppend of their signal vectors
  SVSampleBoth  ::    SVSample svLeft
                   -> SVSample svRight
                   -> SVSample (SVAppend svLeft svRight)
\end{code}
\hrule
\caption{Datatype for signal samples.}
\label{figure:signal_sample_datatype}
\end{figure}

@@

\begin{figure}
\begin{code}
data SVOccurrence sv where
  -- Leaf, carrying a value for the occurrence
  SVOccurrence ::    a
                  -> SVOccurrence (SVEvent a)
  -- Lift an occurrence to be on the left side of a signal vector
  SVOccLeft    ::    SVOccurrence svLeft
                  -> SVOccurrence (SVAppend svLeft svRight)
  -- Lift an occurrence to be on the right side of a signal vector
  SVOccRight   ::    SVOccurrence svRight 
                  -> SVOccurrence (SVAppend svLeft svRight)
\end{code}
\hrule
\caption{Datatype for event occurrences.}
\label{figure:event_occurrence_datatype}
\end{figure}

@@

\begin{figure}
\begin{code}
data SVDelta sv where
  -- A replacement value for a sample of a single signal
  SVDeltaSignal  ::    a
                    -> SVDelta (SVSignal a)
  -- No replacement value for a sample of an arbitrary
  --  signal vector
  SVDeltaNothing ::    SVDelta sv
  -- Combine replacements for component signal vectors into 
  --   replacements for the SVAppend of the vectors.
  SVDeltaBoth    ::    SVDelta svLeft
                    -> SVDelta svRight
                    -> SVDelta (SVAppend svLeft svRight)
\end{code}
\hrule
\caption{Datatype for signal updates.}
\label{figure:signal_update_datatype}
\end{figure}

@@

\begin{figure}
\begin{code}
-- Type tag for signal functions which don't require
--  an initial input
data Initialized

-- Type tag for signal functions which require
--  an initial input
data NonInitialized

-- Type of signal functions
data SF init svIn svOut where
  -- Non-initialized signal functions consist of a function
  --  from an input sample to a pair of an output sample
  --  and an initialized continuation of the signal function
  SF     ::    (SVSample svIn 
                  -> (SVSample svOut,
                      SF Initialized svIn svOut)) 
            -> SF NonInitialized svIn svOut

  -- Initialized signal functions consist of two functions.
  -- The first function's inputs are the time delta, and updates
  --   to the input signals. Its outputs are updates to the output
  --   signals, a set of output events, and a continuation of the
  --   signal function.
  SFInit ::    (Double 
                  -> SVDelta svIn
                  -> (SVDelta svOut,
                      [SVOccurrence svOut],
                      SF Initialized svIn svOut))

  -- The second function's input is an event occurrence for an
  --   event in its input signal vector. Its outputs are a set 
  --   of output occurrences for events in its output signal
  --   vector, and a continuation of the signal function.
            -> (SVOccurrence svIn
                  -> ([SVOccurrence svOut],
                      SF Initialized svIn svOut))
            -> SF Initialized svIn svOut
\end{code}
\hrule
\caption{Datatype and empty types for signal functions.}
\label{figure:signal_function_datatype}
\end{figure}

@@

composeInit ::     SF Initialized svIn svBetween
                -> SF Initialized svBetween svOut
                -> SF Initialized svIn svOut
composeInit (SFInit dtCont1 inputCont1) sf2@(SFInit dtCont2 inputCont2) =
  SFInit
    (\dt sigDelta -> 
       let (sf1MemOutput, sf1EvtOutputs, sf1New) = dtCont1 dt sigDelta
           (sf2MemOutput, sf2EvtOutputs, sf2New) = dtCont2 dt sf1MemOutput
           (sf2EvtEvtOutputs, sf2Newest) = applySF sf2New sf1EvtOutputs
       in (sf2MemOutput,
           sf2EvtOutputs ++ sf2EvtEvtOutputs,
           composeInit sf1New sf2Newest)
    )
    (\evtOcc -> 
      let (sf1Outputs, sf1New) = inputCont1 evtOcc
          (sf2FoldOutputs, sf2New) = applySF sf2 sf1Outputs
      in (sf2FoldOutputs, composeInit sf1New sf2New)   
    )

-- The programmer should not rely on any 
applySF ::    SF Initialized svIn svOut
           -> [SVOccurrence svIn]
           -> ([SVOccurrence svOut],
               SF Initialized svIn svOut)
applySF sf indices =
  foldr (\evtOcc (changes, SFInit _ changeCont) ->
           let (newChanges, sfNew) = changeCont evtOcc
               in (newChanges ++ changes, sfNew))
        ([], sf)
        indices
\end{code}
\hrule
\caption{Implementation of serial composition.}
\label{figure:serial_composition_implementation}
\end{figure}

The {\tt switch} combinator is the means of introducing reactivity into a signal
function. This combinator allows a signal function to replace itself by
producing an event occurrence. The combinator wraps a signal function, and 
observes an event on the right side of the output signal vector. At the first
occurrence of the event, the signal function carried by the occurrence replaces
the signal function.

The {\tt switch} combinator stores the input sample provided during initialization,
and updates it with the input signal updates. When the wrapped signal function
produces an occurrence carrying a new signal function, that signal function is
initialized with the stored input sample. It is then "wrapped" by another
function which closes over its output sample, and outputs the sample as a signal
update as the next time step. After this, it acts as the new signal function.
This wrapping has some performance implications, which are discussed

@@

\begin{figure}
\begin{code}
-- T.swap is imported from Data.Tuple
-- chooseOccurrence, splitSample, splitDelta,
--  combineSamples, combineDeltas, occRight, and
--  occLeft are convenience functions for manipulating
--  signal sample
swap :: (sv1 :^: sv2) :~> (sv2 :^: sv1)
swap =
  SF (\sigSample ->  
       (uncurry combineSamples $ T.swap $ splitSample sigSample, 
        swapInit)

swapInit :: SF Initialized (SVAppend sv1 sv2) (SVAppend sv2 sv1)
swapInit =
  SFInit (\_ sigDelta ->
           (uncurry combineDeltas $ T.swap $ splitDelta sigDelta,
            swapInit) 
         (\evtOcc ->
            (case chooseOccurrence evtOcc of
               Left lOcc  -> [occRight lOcc]
               Right rOcc -> [occLeft rOcc], swapInit))

@@

The {\tt union} combinator's implementation passes along every event occurrence 
it receives on either input, stripping off the left and right combinator. This is 
acceptable since we do not insist on a total ordering of events, or an event time
resolution greater than the sampling rate. The {\tt combineSignals} combinator 
maintains the values of both signals, and applies the combination function whenever
one is updated, producing an update to the output signal's value.
The {\tt capture} combinator similarly maintains the input signal value, and adds it
to each event occurrence.

@@

The implementation strategy leaves room for optimizations. In particular, an
additional constructor for time-independent signal functions would allow
portions of a signal function to forgo evaluation during time steps unless they
had signal updates. Optimizations in the style of Yampa, where an AST for the
current signal function is maintained alongside the continations and checked for
opportunities to eliminate redundant components of the signal function,
might further improve performance. In particular, collapsing nested or
serially-composed switches which have occurred since the last time step would
remove at least some of the observed dependence of performance on sampling rate.
Nevertheless, this implementation performs quite well as it currently exists, as
we demonstrate in Chapter~\ref{chapter:Evaluation_and_Comparisons}.

### 2013-07-17 Docs/Thesis/implementation.tex
thesis updates

Having established the design and semantics of the reactive system, I present a purely 
functional implementation. For this implementation, I will make significant use
of an extension to the Haskell type system. This extension, called
GADTs~\cite{Cheney2003,Xi2003}, provides a mechanism for expressing abstractions
which are not easily expressible in a standard Hindley-Milner type system, while
maintaining the type soundness of Haskell. In particular, GADTs provide the
power we require to take advantage of the signal vector representation of signal
function inputs and outputs.

>>>>

Having established the design and semantics of the reactive system, I present a purely 
functional implementation. For this implementation, I will make significant use of
Generalized Algebraic Datatypes (GADTs)~\cite{Cheney2003,Xi2003}. GADTs provide a mechanism
for expressing abstractions which are not easily expressible in a standard
Hindley-Milner type system, while maintaining the type soundness of Haskell.
In particular, GADTs provide the power we require to take advantage of the
signal vector representation of signal function inputs and outputs.

### 2013-07-17 Docs/Thesis/system_design_and_interface.tex
thesis updates

-- Make right-associative
associate :: ((sv1 :^: sv2) :^: sv3) :~> (sv1 :^: (sv2 :^: sv3))

-- Make left-associative
unassociate :: (sv1 :^: (sv2 :^: sv3)) :~> ((sv1 :^: sv2) :^: sv3)
\end{code}
\hrule
\caption{Routing combinators.}
\label{figure:routing_combinators}
\end{figure}

@@

The {\tt combineSignals} combinator applies a binary function pointwise to two
signals, and produces the result of this application as a third signal. The
combining function is necessary because we will have two input samples at each
time step, and must produce exactly one output sample.

>>>>

-- Make right-associative
associateRight :: ((sv1 :^: sv2) :^: sv3) :~> (sv1 :^: (sv2 :^: sv3))

-- Make left-associative
associateLeft :: (sv1 :^: (sv2 :^: sv3)) :~> ((sv1 :^: sv2) :^: sv3)
\end{code}
\hrule
\caption{Routing combinators.}
\label{figure:routing_combinators}
\end{figure}

@@

The {\tt combineSignals} combinator applies a binary function pointwise to two
signals, and produces the result of this application as a third signal. The
combining function is necessary because we will have two input samples at each
time step, and must produce exactly one output sample. Even if we restrict the types
to be the same, parametricity forbids assuming that we can combine two values of
an arbitrary type.\footnote{The {\tt union} combinator does not encounter this difficulty
since event occurrences, being temporally discrete, can be temporally interleaved without
 combining their individual values}
